Revisiting the impact of deepfake technology towards viewer's political choice: A replication of Soubhik Barari (2021)

### 2021-03-22 outputs/paper/paper.Rmd
Update Discussion and Abstract

---
title: "Revisiting the impact of deepfake technology towards viewer's political choice: A replication of Soubhik Barari (2021)"
author: "Yingying Zhou, Yang Wu, Xinyi Xu, Hong Pan"
output:
 bookdown::pdf_document2:
 toc: no
abstract: "We reproduce Soubhik Barari (2021) paper and identify several additional demographic traits contributing to the persuasion power of deepfake on political defamation through its material effect on feeling manipulation in a RCT field experiment on the 2020 Presidential election. We find little evidence that deepfake videos have a unique ability to shift viewers’ perceptions on politicians. Apart from partisanship, education, and other socioeconomic factors identified by the original paper, we discover that income and internet usage would also significantly impact the candidate’s affective appeal to the news feed viewer. "

@@

[@citePaper] examines the news feed intervention through two Randomized Controlled Trial field experiments on the 2020 U.S. Democratic presidential election to investigate the persuasion power of deepfake. In the first experiment, [@citePaper]exposes participants to one of five different forms of false media treatment including deepfaked videos and compares viewer’s deceptive level and affective response. In the second experiment, [@citePaper] measures the viewer's ability to discriminate between a feed of 8 real and fake videos.

@@

# Data
## Dataset features
The paper intends to replicate the experimental report by [@citePaper], Christopher Lucas, and Kevin Munger. We are going to analyze the dataset collected by their experiment. The survey dataset records 5,750 observations with 100 variables. In this case, we focus on the top 20 most important features related to the favorability towards the Democratic politician Elizabeth Warren. It covers viewer traits on demographic, socioeconomic status, and political related positions that are immutable or intervenable characteristics to deepfakes' misinformation. The authors hypothesized several risky subgroups that may be differentially susceptible to deepfakes. That included the following categories of media consumers:

@@

This paper replicates the first experiment of Soubhik Barari (2021) to further study the persuasion power of deepfake through the channel of its affective manipulation on viewer’s political perception. Media heterogeneity effects are reviewed via t-tests between other false media and deepfake.  The heterogeneity in affective response by viewer traits is examined with a multiple linear regression model. Feature selection is performed via the Random Forest classifier to select the top 20 relevant variables into our model input, which identified additional important socioeconomic features other than the ones from the original paper.

@@

Some modelling bias and ethical concerns are addressed:

### Privacy problem on zip code
To protect the privacy of respondents in the experiment, detailed address information such as zip code was excluded from the linear regression model. Instead, location was proxied by region. Thereby, it would be impossible to reverse engineer the identity of the respondent by any ill-intended third parties. Furthermore, postal code is a proxy for socioeconomic status, which correlates with HHI (household income), causing multicollinearity and eventually biasing the coefficient [@citeCorrelation].

### Choice of candidate - election result disruption & pre-existing gender prejudice
The one-month experiment was run prior to the 2020 U.S. presidential election, to minimize the risk of influencing the election outcome and in the meantime to have discernible effect for the experiment, a high-profile candidate who was not finalisted to be the democratic nominee should be selected. In addition, we would like to test for the impact of pre-existing sexism against women among subjects on the effect of the deepfake. Elizabeth Warren came naturally to be our choice of the candidate who meets all the conditions above.

### Active debrief - post-experiment potential change in participants’ political behavior
At the completion of the survey, an extensive debrief was conducted on all subjects which instructed respondents to affirm that the treatment media about Elizabeth Warren was false. The active debrief was done to educate participants on deepfake media. And thus it would be unlikely for the treated group to change their political behavior later on as a post-exposure consequence.

@@

### Heterogeneity in different socioeconomic groups
As for the regression model result on affective response towards Elizabeth Warren, heterogeneous effects in respondents from different socioeconomic groups were observed. All else held constant, on average people with a postgraduate degree significantly favor Warren by 9.3 points more than people with lower than high school education. High-incomers with annual income above $100k favor Warren by 4 points more than people who earn less than $25,000 a year. In contrast, people who were unwilling to disclose their salaries disliked the candidate by 4.94 points compared to the low-incomers. Elderly people aged over 65 deduct feeling marks by 4.35 points towards Warren after being exposed to the deepfake media about Warren.

### Impact from partisan motivated reasoning and sexism discrimination
Partisanship accounts for a large part of the sentiment score towards Warren. For people supporting rival parties to the Democratic as the Republican and Independent, they substantially dislike Warren by about 40 and 27 points on average respectively compared with people identifying with the Democratic. Additionally, the map would also justify the positive coefficient for the Northeast region, where Warren is from; she is home based in Massachusetts [@citeWarrenBase]. For people holding ambivalent sexism views against women, a higher level in their hostile attitude would deduct 4 more marks from their feeling scales for the female candidate Warren on average.

### Deepfake and other false media, not that different in fooling the public
Although there is significant influence from the deepfake video on altering people’s sentiment towards the candidate, the two-sample t-test results between deepfake and other false media assure us that the its degree of detriment is no different from false audio or text in terms of deception and affective impression.

### Political knowledge and internet usage
It is worth our attention that compared to political knowledge, internet usage has a more significant effect on respondents’ affective impression. Perhaps that the more frequent interest users gain a more holistic understanding of political positions and current trends in the world thus they are more immune to the infamatory deepfake technologies.

@@

**Democratic party only**: The paper chose the candidate to be a politician in the Democratic party, and focuses the topic on the Democratic election. It is possible that the experiment outcome was only valid for the analysis on Democratic party, especially supporter traits on socioeconomic and demographic characteristics.

@@

**Time window**: for the experiment time window, it can be prolonged or randomized in a series of experiment repetitions as long as they are conducted after the time point when the in-party presidential nominee has been finalized.

@@

**Additional features**: Apart from the treatment variable and regressors describing demographic and socioeconomic traits (age, education, gender, party identification, political knowledge, metrics on internet usage, .etc) in the original model, the random forest algorithm identified household income (`HHI`), region, script type, and ethnicity as relevant features. Income confounds with deepfake deception and affective appeal through its correlation with digital literacy and internet usage. Region and ethnicity are correlated with partisanship. The five types of script contents were included to control for any differential effect of persuasion ability by deepfake.

>>>>

---
title: "Revisiting the impact of deepfake technology towards viewer's political choice: A replication of Soubhik Barari (2021)"
author: "Yingying Zhou, Yang Wu, Xinyi Xu, Hong Pan"
output:
 bookdown::pdf_document2:
 toc: no
abstract: "We reproduce Soubhik Barari (2021) paper and identify several additional demographic traits contributing to the persuasion power of deepfake on political defamation through its material effect on feeling manipulation in a RCT field experiment on the 2020 Presidential election. With the growing ubiquity of deepfake technology, here comes one more channel for the spread of misinformation. Despite it being a vicious threat, we find little evidence that deepfake videos have a unique ability to shift viewers’ perceptions on politicians. Apart from partisanship, education, and other socioeconomic factors identified by the original paper, we discover that income and internet usage would also significantly impact the candidate’s affective appeal to the news feed viewer. "

@@

[@citePaper] examines the news feed intervention through two Randomized Controlled Trial field experiments on the 2020 U.S. Democratic presidential election to investigate the persuasion power of deepfake. In the first experiment, [@citePaper] exposes participants to one of five different forms of false media treatment including deepfaked videos and compares viewer’s deceptive level and affective response. In the second experiment, [@citePaper] measures the viewer's ability to discriminate between a feed of 8 real and fake videos.

@@

# Data
## Dataset features
The paper intends to replicate the experimental report by [@citePaper]. We are going to analyze the dataset collected by their experiment. The survey dataset records 5,750 observations with 100 variables. In this case, we focus on the top 20 most important features related to the favorability towards the Democratic politician Elizabeth Warren. It covers viewer traits on demographic, socioeconomic status, and political related positions that are immutable or intervenable characteristics to deepfakes' misinformation. The authors hypothesized several risky subgroups that may be differentially susceptible to deepfakes. That included the following categories of media consumers:

@@

This paper replicates the first experiment of [@citePaper] to further study the persuasion power of deepfake through the channel of its affective manipulation on viewer’s political perception. Media heterogeneity effects are reviewed via t-tests between other false media and deepfake.  The heterogeneity in affective response by viewer traits is examined with a multiple linear regression model. Feature selection is performed via the Random Forest classifier to select the top 20 relevant variables into our model input, which identifies additional important socioeconomic features such as household income and region other than the ones from the original paper.

@@

Some modelling bias and ethical concerns in our model and the original paper are addressed:

### Privacy problem on zip code
To protect the privacy of respondents in the experiment, detailed address information such as zip code was excluded from our linear regression model. Instead, location was proxied by region. Thereby, it would be impossible to reverse engineer the identity of the respondent by any ill-intended third parties. Furthermore, postal code is a proxy for socioeconomic status, which correlates with HHI (household income), causing multicollinearity and eventually biasing the coefficient [@citeCorrelation].

### Choice of candidate - election result disruption & pre-existing gender prejudice
In the original paper, the one-month experiment was run prior to the 2020 U.S. presidential election, to minimize the risk of influencing the election outcome and in the meantime to have discernible effect for the experiment, a high-profile candidate who was not finalisted to be the democratic nominee should be selected. In addition, [@citePaper] would like to test for the impact of pre-existing sexism against women among subjects on the effect of the deepfake. Elizabeth Warren came naturally to be the choice of the candidate in the [@citePaper] paper who meets all the conditions above.

### Active debrief - post-experiment potential change in participants’ political behavior
At the completion of the survey by [@citePaper], an extensive debrief was conducted on all subjects which instructed respondents to affirm that the treatment media about Elizabeth Warren was false. The active debrief was done to educate participants on deepfake media. And thus it would be unlikely for the treated group to change their political behavior later on as a post-exposure consequence.

@@

### Heterogeneity in different socioeconomic groups
As for the regression model result on affective response towards Elizabeth Warren, heterogeneous effects in respondents from different socioeconomic groups were observed, which are in line with the findings by [@citePaper]. All else held constant, on average people with a postgraduate degree significantly favor Warren by 9.3 points more than people with lower than high school education. High-incomers with annual income above $100k favor Warren by 4 points more than people who earn less than $25,000 a year. In contrast, people who were unwilling to disclose their salaries disliked the candidate by 4.94 points compared to the low-incomers. Elderly people aged over 65 deduct feeling marks by 4.35 points towards Warren after being exposed to the deepfake media about Warren. [@citePaper] finds that on average postgraduates significantly favor Warren by 12.71 points more and elderly people dislike Warren more by 5.31 points than people younger than 65. The difference is that we include a new feature of income (HHI) which shares some of the variation in affective level with education compared with what [@citePaper] finds.

### Impact from partisan motivated reasoning and sexism discrimination
Partisanship accounts for a large part of the sentiment score towards Warren. For people supporting rival parties to the Democratic as the Republican and Independent, they substantially dislike Warren by about 40 and 27 points on average respectively compared with people identifying with the Democratic. The coefficient results echo with [@citePaper]’s result of -39 and -27.5. Additionally, the map would also justify the positive coefficient for the Northeast region, where Warren is from; she is home based in Massachusetts [@citeWarrenBase]. The effect of regions is not accounted for by [@citePaper]’s paper. For people holding ambivalent sexism views against women, a higher level in their hostile attitude would deduct 4 more marks from their feeling scales for the female candidate Warren on average, roughly the same as -3.6 by [@citePaper]’s result.

### Deepfake and other false media, not that different in fooling the public
Although there is significant influence from the deepfake video on altering people’s sentiment towards the candidate, the two-sample t-test results between deepfake and other false media assure us that the its degree of detriment is no different from false audio or text in terms of deception and affective impression. Our conclusion is consistent with [@citePaper]’s.

### Political knowledge and internet usage
It is worth our attention that compared to political knowledge, internet usage has a more significant effect on respondents’ affective impression. On average, one unit increase in interest usage frequency will increase viewer’s affective score on Warren by 1.12, matching with [@citePaper]’s 1.30.  Perhaps that the more frequent interest users gain a more holistic understanding of political positions and current trends in the world thus they are more immune to the infamatory deepfake technologies.

@@

**Democratic party only**: The paper follows [@citePaper] to choose the candidate to be a politician in the Democratic party, and focuses the topic on the Democratic election. It is possible that the experiment outcome was only valid for the analysis on Democratic party, especially supporter traits on socioeconomic and demographic characteristics.

@@

**Time window**: For the experiment time window, it can be prolonged or randomized in a series of experiment repetitions as long as they are conducted after the time point when the in-party presidential nominee has been finalized.

@@

**Additional features**: Apart from the treatment variable and regressors describing demographic and socioeconomic traits (age, education, gender, party identification, political knowledge, metrics on internet usage, .etc) in the original paper, the random forest algorithm identified household income (`HHI`), region, script type, and ethnicity as relevant features. Income confounds with deepfake deception and affective appeal through its correlation with digital literacy and internet usage. Region and ethnicity are correlated with partisanship. The five types of script contents were included to control for any differential effect of persuasion ability by deepfake.
